---
title: "DS 3001 Final Project: Predicting Attitudes Towards Abortion"
author: "Maddie Ashby, Claire Dozier, Emma Murphy, Kylie Wise"
date: "12/6/2021"
output:
  html_document:
    number_sections: yes
    toc: yes
    toc_float: yes
    code_folding: hide
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Question and Background Information

## Question and Context



## Data Background Information 




## 



# Data Preparation and Cleaning

```{r Load libraries, include=FALSE}
library(readxl)
library(dplyr)
library(tidyverse)
library(gtsummary)
library(ggplot2)
library(randomForest)
library(caret)
library(rio)

library(rpart)
library(psych)
library(pROC)
library(rpart.plot)
library(rattle)
library(caret)
library(C50)
library(mlbench)

```


```{r warning = FALSE, console = FALSE, message=FALSE}

# Read in data 
SurveyResults <- read_excel("SurveyResults.xlsx")

# Subset relevant columns
data <- SurveyResults[,c(3:10,17,18,20,21)]

# Rename columns
colnames(data) <- c("country","gender","age","employment","edu","marital_status","household_income","chief_income_earner","close_lgbtq","attitude","had_abortion","religiosity")

# set Na to NA
data[data=="Na"] <- NA

sapply(data, function(x) sum(is.na(x)))

# remove all NA's
#data <- na.omit(data)

# Recode NAs
data <- data %>% replace_na(list(employment = "Employment_Status_Unavailable", marital_status = "Marital_Status_Unavailable", household_income = "Income_Unavailable"))

#sapply(data, function(x) sum(is.na(x)))

# Remove observations where attitude = don't know
data <- data[!(data$attitude == "Don't know/Prefer not to say"),]

# Factor collapsing 
data$employment<-fct_collapse(data$employment,
                              student = "A student",
                              employed = c("Full time", "Part time", "Self employed"),
                              unemployed = c("Retired", "Unemployed", "Prefer not to answer"),
                              other = c("Employment_Status_Unavailable")
                              )

# unique(data$employment), check that factor worked

data$attitude<-fct_collapse(data$attitude,
legal = c("Abortion SHOULD be permitted whenever a woman decides she wants one","Abortion SHOULD be permitted in certain circumstances, such as if a woman has been raped"),
illegal = c("Abortion should NOT be permitted under any circumstances, except when the life of the mother is in danger", "Abortion should NEVER be permitted, no matter what circumstance exists"))

#data$attitude <- recode(data$attitude, "legal" = "1", "illegal" = "0")
#data$attitude <- as.factor(data$attitude)

# unique(data$attitude), check that it worked 

data$marital_status <- fct_collapse(data$marital_status,
  married = c("Married", "Domestic partnership / Living as married"),
  single = c("Single"),
  other = c("Divorced", "Widowed", "Marital_Status_Unavailable"))

# unique(data$marital_status), check that it worked 

data$country <- fct_collapse(data$country,
  americas = c("United States", "Canada", "Argentina", "Brazil", "Mexico"), #1883
  asia = c("India", "Japan", "South Korea", "China", "Russia"), #3477
  europe = c("Great Britain", "Ireland", "Italy", "Hungary", "Belgium", "Sweden", "Spain", "Poland", "Turkey", "Germany", "France"), #5193
  other = c("South Africa", "Australia")) #1501

# unique(data$country)

data$had_abortion <- fct_collapse(data$had_abortion,
  No = "No",
  Yes = "Yes",
  Other = c("Not sure", "Prefer not to answer"))

# unique(data$had_abortion), make sure that it worked 

data$religiosity<-fct_collapse(data$religiosity,
                               often = c("Every week", "More than once a week", "Nearly every week"),
                               occasionally = c("Once a month", "Two to three times a month"),
                               rarely = c("Several times a year", "Once a year", "Less than once a year"),
                               never = c("Never", "Don’t know/no answer")
                               )

# Update additional variables to factors
data$gender <- as.factor(data$gender)
data$edu <- as.factor(data$edu)
data$household_income <- as.factor(data$household_income)
data$chief_income_earner <- as.factor(data$chief_income_earner)
data$close_lgbtq <- as.factor(data$close_lgbtq)

# unique(data$religiosity) # make sure that it worked 
str(data)
```


# Exploratory Data Analysis 

## Calculate Prevalence 

## High-Level Summary of Variables
```{r}
data %>% tbl_summary()
```

```{r}
# CLAIRE'S EDA

# histogram of age distribution for survey
hist(data$age)

# household income
income_count <- data %>% count(household_income)

ggplot(income_count, aes(x = household_income, y=n)) + 
  geom_bar(stat = "identity")

# abortion by employment

ab_edu <- data %>% group_by(edu) %>% count(had_abortion)
ggplot(ab_edu, aes(x = edu, y = n, fill = had_abortion)) +
  geom_col(position = "dodge")


# abortion by marriage
ab_marriage <- data %>% group_by(marital_status) %>% count(had_abortion)
ggplot(ab_marriage, aes(x = marital_status, y = n, fill = had_abortion)) +
  geom_col(position = "dodge")

# abortion and age
ab_age <- data %>% group_by(age) %>% count(had_abortion)
ab_age <- ab_age[ab_age$had_abortion == "Yes",]
ggplot(ab_age, aes(x=as.factor(age), y = n)) +
  geom_bar(stat="identity")

```

```{r}
# EMMA'S EDA
counts_gender <- table(data$gender, data$attitude)
barplot(counts_gender, main="Attitude by Gender",
        xlab="Gender", col=c("darkblue","red"),
        legend = rownames(counts_gender), beside=TRUE)

counts_country <- table(data$country, data$attitude)
barplot(counts_country, main="Attitude by Country",
        xlab="Country", col=c("darkblue","red", "green", "yellow"),
        legend = rownames(counts_country), beside=TRUE)

counts_education <- table(data$edu, data$attitude)
barplot(counts_education, main="Attitude by Education",
        xlab="Education", col=c("darkblue","red", "green"),
        legend = rownames(counts_education), beside=TRUE)

counts_abortion <- table(data$had_abortion, data$attitude)
barplot(counts_abortion, main="Attitude by Whether or Not The Respondent or Someone Close to Respondent has had an Abortion",
        xlab="Respondent or Someone Close to Respondent has had an Abortion",
        col=c("darkblue","red", "green"),
        legend = rownames(counts_abortion), beside=TRUE)

counts_income <- table(data$household_income, data$attitude)
barplot(counts_income, main="Attitude by Household Income",
        xlab="Household Income", col=c("darkblue","red", "green", "yellow"),
        legend = rownames(counts_income), beside=TRUE)

counts_breadwinner <- table(data$chief_income_earner, data$attitude)
barplot(counts_breadwinner, main="Attitude by Whether or Not The Respondent is Breadwinner",
        xlab="Whether or Not The Respondent is Breadwinner", col=c("darkblue","red"),
        legend = rownames(counts_breadwinner), beside=TRUE)

counts_religiosity <- table(data$religiosity, data$attitude)
barplot(counts_religiosity, main="Attitude by Whether or Not The Respondent is Religious",
        xlab="Whether or Not The Respondent is Religious", col=c("darkblue","red", "green", "yellow"),
        legend = rownames(counts_religiosity), beside=TRUE)
```

```{r}
# MADDIE's EDA
ggplot(data, aes(x=gender, fill=gender)) + 
  geom_bar() + 
  facet_wrap(~attitude)

ggplot(data, aes(x=age)) +
  geom_bar(colour="salmon") +
  facet_wrap(~attitude)

ggplot(data, aes(x=age, fill=gender)) +
  geom_bar() +
  facet_wrap(~attitude)
```

# Methods: Model Building and Evaluation

```{r}
# Calculate the prevalence 
target_table <- table(data$attitude)

baserate <- target_table[[2]]/(target_table[[1]] + target_table[[2]])
baserate

```


## Partition the Data

```{r Partition the data}

set.seed(2000)
part_index_1 <- caret::createDataPartition(data$attitude,
                                           times=1,
                                           p = 0.70,
                                           groups=1,
                                           list=FALSE)

train <- data[part_index_1, ]
tune_and_test <- data[-part_index_1, ]


tune_and_test_index <- caret::createDataPartition(tune_and_test$attitude,
                                           p = .5,
                                           list = FALSE,
                                           times = 1)

tune <- tune_and_test[tune_and_test_index, ]
test <- tune_and_test[-tune_and_test_index, ]

dim(train)
dim(tune)
dim(test)
```

## Build and Evaluate Initial Decision Tree Model

```{r warning=FALSE, message=FALSE, console=FALSE}
# Choose the features and classes
features <- train[,c(-10)]
target <- train$attitude

str(features)
str(target)

# Cross validation process 
fitControl <- trainControl(method = "repeatedcv",
                          number = 10,
                          repeats = 5, 
                          returnResamp="all",
                          classProbs = TRUE,
                          allowParallel = TRUE) 

grid <- expand.grid(.winnow = c(TRUE,FALSE), 
                    .trials=c(1,5,10,15,20), 
                    .model="tree")

set.seed(2000)
abortion_dt_mdl <- train(x=features,
                y=target,
                method="C5.0",
                tuneGrid=grid,
                trControl=fitControl,
                verbose=TRUE)
```

```{r}
abortion_dt_mdl
```
```{r}
# visualize the re-sample distributions
xyplot(abortion_dt_mdl,type = c("g", "p", "smooth"))
varImp(abortion_dt_mdl)
```

```{r}
# predict performance 
abort_dt_pred_tune = predict(abortion_dt_mdl, tune, type= "raw")

abort_dt_eval <- confusionMatrix(as.factor(abort_dt_pred_tune), 
                as.factor(tune$attitude), 
                dnn=c("Prediction", "Actual"), 
                mode = "everything")

table(tune$attitude)
#(wine_pred_tune_p = predict(wine_mdl,tune,type= "prob"))
abort_dt_eval 
```

```{r}

```



## Build and Evalaute Initial Random Forest Ensemble Model
```{r}
mtry_tune <- function(x){
  xx <- dim(x)[2]-1
  sqrt(xx)
}

mtry_tune(train)

set.seed(1984)	
abortion_RF_mdl = randomForest(attitude~.,          
                            train,     
                            #y = NULL,           
                            #subset = NULL,      
                            #xtest = NULL,       
                            #ytest = NULL,       
                            ntree = 100,  # 500 tree       
                            mtry = 7,  # mtry = 3 (calculated above)          
                            replace = TRUE,      
                            #classwt = NULL,     
                            #strata = NULL,      
                            sampsize = 200,      
                            nodesize = 5,        
                            #maxnodes = NULL,    
                            importance = TRUE,   
                            #localImp = FALSE,   
                            norm.votes = TRUE,   
                            do.trace = TRUE,     
                            keep.forest = TRUE,  
                            keep.inbag = TRUE)  
```
```{r}
err.rate_mdl1 <- as.data.frame(abortion_RF_mdl$err.rate)
err.rate_mdl1
```


```{r}
abortion_RF_mdl
```


## Tune the RF Model


## Try an XGBoost model

```{r}

```


# Final Model Evaluation 



# Conclusions




# Future Work



